Possible Inconsistency in Maximum Likelihood Calculation of Ancestral States

### 2017-03-21 joint_inf.tex
Some intro and discussion.

@@

\subsection{The usual story}

@@

>>>>

\section*{Introduction}

Classical maximum-likelihood estimation in phylogenetics operates by integrating out possible ancestral states at the internal nodes of the tree.
Recently Neher et al have suggested using an approximation to ML inference in which likelihood is maximized jointly across model parameters and ancestral sequences.
This is attractive from a computational perspective, because joint ML inference can proceed according to an EM-like algorithm in which ML ancestral sequences are first inferred and then model parameters are optimized conditioned on the ancestral sequences.
This latter step is much simpler and more computationally efficient when conditioning on ancestral sequences compared to the case in which ancestral sequences are marginalized out.

However, existing consistency proofs for phylogenetics [cite] do not apply to the setting of joint inference of trees and ancestral sequences, raising the question of if this procedure enjoys good statistical properties.
There are examples in statistics, such as the Neyman-Scott paradox \cite{Neyman1948-tt}, in which additional parameters destroy consistency.

In this paper we show that the joint inference of trees and ancestral sequences is not consistent.
To do so, we follow a similar strategy of using simple models and four taxon trees.
Here we use a binary symmetric model and simulate on the ``Farris zone'' \cite{Siddall1998-hq} tree.
We use bounds on likelihood to demarcate a sizeable area of long branch lengths in which joint inference gives the wrong tree.

@@

\subsection{Classical maximum-likelihood inference in phylogenetics}

@@

\section{Discussion}

Neyman-Scott paradox.

Interesting that here we are simulating on the Farris tree and end up with the Felsenstein tree.
For maximum parsimony it's the opposite \cite{Felsenstein1978-rr}.
Discussion of number of parameters of each.

However, note that \cite{Siddall1998-hq} get things going the same way, although \cite{Swofford2001-hr} show that the problem is that they didn't simulate long enough sequences.
